---
title: 'Assignment #1 - Data Wrangling in R'
author: "François d'Entremont"
date: '2022-10-19'
output:
  pdf_document: default
  default: default
  html_document:
    df_print: paged
---

First let's load all the packages needed for the assignment

```{r}
library(tidyverse)
library(sf)
library(tmap)
library(sp)
```

Now let's load the data and take a look at it. The result is a tibble so we can
use our libraries on them.
```{r}
# load world data using csv file
world_data <- read_csv("world_data.csv")
View(world_data)
class(world_data)
```

# Task #1 - Subsetting Rows and Columns

We are looking for countries located in Eastern Europe with male life expectancy
greater than 68 years and have negative population growth.

```{r}
world_data_subset_piping <- world_data %>% filter(
    Region == "East Europe", `Life Expectancy (Male)` > 68,
    `Population Increase (% per year)` < 0) %>% select(
      Country, Region, `Life Expectancy (Male)`,
      `Population Increase (% per year)`)

View(world_data_subset_piping)
world_data_subset_piping
```

The chunk of code above pipes the data into the filter function and then pipes
that data into a select function to display the relevant columns.

# Task #2 - Summary Statistics at the Regional Scale

Calculate the male to female life expectancy ratio and output a grouped summary
of the average male to female life expectancy at the regional scale.

```{r}
world_data_m2fer <- world_data %>% group_by(Region) %>% summarise(
  `Male to Female Life Expectancy Ratio` = mean(
    `Life Expectancy (Male)` / `Life Expectancy (Female)`)) 

View(world_data_m2fer)
world_data_m2fer
```

# Task #3 - Mapping Spatial Data in R

Loading the world dataset from the shapefile and viewing it
```{r}
world <- st_read("world.shp")

View(world)
class(world)
```
We see that the column with the country names doesn't match with the column
Country in world_data so we'll change it so we can merge them together. We also
see that it's a data.frame so we can't use our packages on it.

```{r}
# We must give the columns the same name before we join them
names(world)[1] = "Country"

View(world)
```

Let's check if the entries in the country columns match.
```{r}
sum(world$Country == world_data$Country) == 108
```
All 108 entries match!

Now let's do the join! We can join a data.frame to a tibble using the inner_join
function and the result will be a tibble.
```{r}
world_join <- inner_join(world_data, world)
View(world_join)
class(world_join)
```
Removing all the countries with male to female literacy rates and daily caloric
intake values of zero with a single filter function

```{r}
world_join_filter <- filter(world_join, `Literacy Rate for Males (%)` != 0,
                            `Literacy Rate for Females (%)` != 0,
                            `Daily Calorie Intake` != 0)

View(world_join_filter)
```

We see that there are now 59 entries. We removed almost half the countries.

We will now calculate the population density for the remaining countries. We
will add a column called Population Density. This column will store values in
persons per square kilometre

```{r}
world_join_filter$`Population Density` <- 
  world_join_filter$Population / world_join_filter$Area

View(select(world_join_filter, Country, `Population Density`))
```

We will now calculate the male to female literacy ratio for the remaining 
countries. We will add a column called Literacy Ratio.

```{r}
world_join_filter$`Literacy Ratio` <- 
  world_join_filter$`Literacy Rate for Males (%)` / 
  world_join_filter$`Literacy Rate for Females (%)`

View(select(world_join_filter, Country, `Literacy Rate for Males (%)`, 
            `Literacy Rate for Females (%)`, `Literacy Ratio`))
```

Now we need to find the top 5 countries in terms of population density, in 
descending order.

```{r}
world_join_top5density <- select(head(arrange(
  world_join_filter, desc(`Population Density`)),5),
  Country, `Population Density`)

world_join_top5density
```

Now we need to find the bottom 5 countries in terms of literacy ratio, in
descending order.

```{r}
world_join_bot5m2flr <- select(head(arrange(
  world_join_filter, desc(`Literacy Ratio`)),5), Country, `Literacy Ratio`)

world_join_bot5m2flr
```

Now we will produce two choropleth graphs showing Population Density and Male 
to Female Literacy

Checking the class when using the st_as_sf function on our data

```{r}
class(st_as_sf(world_join_filter))
```
We must convert our dataset to an sf object to create the graphs.

Below are two choropleth maps with one showing Male to Female Literacy ratio and the other
showing Population Density.

```{r}
M2F_Literacy_Ratio_Graph <- tm_shape(st_as_sf(world_join_filter)) +
  tm_fill("Literacy Ratio", title = "Male to Female Literacy Ratio", 
          palette = c("#c8d4ef", "#94a3d2", "#6573b3", "#3a4593", "#051971")) +
  tm_borders() + tm_layout("World Map")

M2F_Literacy_Ratio_Graph
```


```{r}
Population_Density_Graph <- tm_shape(st_as_sf(world_join_filter)) +
  tm_fill("Population Density", title = "Population Density",
          breaks = c(0, 50, 200, 800, 6000),
          palette = c("#d5e8dd", "#9bd4b2", "#45a36b", "#0b4a25")) +
  tm_borders() + tm_layout("World Map")

Population_Density_Graph
```





